Unsupervised embeddings of organizations capture latent structure of scientific mobility

### 2020-03-08 paper/current/main.tex
removed the \SGNS macro

% CUSTOM MACROS
\def\SGNS{\textit{SGNS}}

@@

This finding relates to the objective function of the \SGNS{} model. The model attempts to maximize the average log probability of observing a target word in a sentence, $w_{t}$, given the words in the surrounding context window, $w_{t + k}$. This can be written as follows, where $c$ is the size of the context window and $T$ iterates through possible target words:

@@

Hence, the \SGNS{} model can be an effective model for investigating any kind of mobility trajectory data, stemming from the mathematical relationship between it and the gravity model of mobility.

@@

In addition to learning geography, its clear that the \SGNS{} model is also encoding the latent prestige hierarchies of academic mobility.
To test this, we derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings.
SemRank is defined from the position of organizations along the prestige axis, with the top ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.81$, Fig.~\ref{fig:semaxis}b).
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$).
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking mean-normalized citation score ($\rho = 0.87$, Fig.~\ref{fig:supp:times_vs_leiden}).
This suggests that \SGNS{} model, in addition to learning the latent geography of mobility, while capture more nuanced characteristics such as latent hierarchy in the axis of the embedding, which is fascinating characteristic of our approach.

@@

By learning a distance directly from the data, the \SGNS{} model successfully  and efficiently captures a robust measure of semantic distance between entities, all with existing tools and minor data processing.
This methodology is broadly applicable to studies of mobility.
Its success also points to the broad applicability of ideas from language and graph embedding techniques such as GLoVE~\autocite{pennington2014glove}, BERT~\autocite{devlin2018bert}, hyperbolic embeddings~\autocite{nickel2017poincare, linzhuo2020hyperbolic}, and persona2vec () to other problem domains, such as the social sciences.
In addition to their technical advantages and the unique suite of methodologies they allow, they are also  readily transferable and can be trivially fit into a diverse array of analyses.

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling (\SGNS) model, for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} .
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance.
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory.
Vectors are learned by training a neural network to optimize the prediction of so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory.
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The \SGNS{} model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window  considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating analysis for each combination.
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the \SGNS{} training for five epochs over all mobility sentences.
All word embeddings are trained using the \textit{gensim} python implementation of the \SGNS{} \textit{word2vec} model.
We visualize embeddings by projecting into two dimensions using UMAP~\autocite{mcinnes2018umap}.

>>>>

% CUSTOM MACROS

@@

This finding relates to the objective function of the skip-gram model. The model attempts to maximize the average log probability of observing a target word in a sentence, $w_{t}$, given the words in the surrounding context window, $w_{t + k}$. This can be written as follows, where $c$ is the size of the context window and $T$ iterates through possible target words:

@@

Hence, the skip-gram word embedding model can be an effective model for investigating any kind of mobility trajectory data, stemming from the mathematical relationship between it and the gravity model of mobility.

@@

In addition to learning geography, its clear that the word embedding also encodes the latent prestige hierarchies of academic mobility.
To test this, we derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings.
SemRank is defined from the position of organizations along the prestige axis, with the top ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.81$, Fig.~\ref{fig:semaxis}b).
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$).
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking mean-normalized citation score ($\rho = 0.87$, Fig.~\ref{fig:supp:times_vs_leiden}).
This suggests that word embeddings, in addition to learning the latent geography of mobility, while capture more nuanced characteristics such as latent hierarchy in the axis of the embedding, which is fascinating characteristic of our approach.

@@

By learning a distance directly from the data, word embeddings successfully  and efficiently captures a robust measure of semantic distance between entities, all with existing tools and minor data processing.
This methodology is broadly applicable to studies of mobility.
Its success also points to the broad applicability of ideas from language and graph embedding techniques such as GLoVE~\autocite{pennington2014glove}, BERT~\autocite{devlin2018bert}, hyperbolic embeddings~\autocite{nickel2017poincare, linzhuo2020hyperbolic}, and persona2vec () to other problem domains, such as the social sciences.
In addition to their technical advantages and the unique suite of methodologies they allow, they are also  readily transferable and can be trivially fit into a diverse array of analyses.

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling model, for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} .
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance.
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory.
Vectors are learned by training a neural network to optimize the prediction of so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory.
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The skip-gram model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window  considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating analysis for each combination.
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the skip-gram training for five epochs over all mobility sentences.
All word embeddings are trained using the \textit{gensim} python implementation of the skip-gram \textit{word2vec} model.
We visualize embeddings by projecting into two dimensions using UMAP~\autocite{mcinnes2018umap}.
